Sperm do not evolve to collaborate in female meiotic drive.

### 2014-05-09 writeup_and_notes/male_dependent_drive.tex
edited male geno. text

%I SUBSTANTIALLY REWROTE THIS
\yb{
The positive frequency dependence of a self-promoting driver can
induce bistability in its spread.
%OLD GRAHAM WORDS
 %, i.e. there is a critical frequency
%above which it can spread quickly to fixation and below which it moves towards loss (an
%unstable equilibrium frequency). In 
%the appendix, we develop a simple
%approximation to this bistability. 
When $s$ is too large to allow a rare driver to increase in frequency, but small enough to allow a common driver to fix (the inequality in EQUATION XXX),
there is  an unstable equilibrium frequency (approximated in EQUATION XXX)  
	above which it can spread quickly to fixation and below which it moves towards loss 
	(see Figure \ref{Bistab_homozyg_cost_fig} for exact results).	
%OLD GRAHAM WORDS
%This bistability occurs in the region of parameter space in Figure
%\ref{Invasion_space}. In Supplementary Figure
%\ref{Bistab_homozyg_cost_fig} we further explore this bistability and
%show that the unstable equilibrium frequency is high across much of
%this parameter space, and so in practice the self promoting allele
%only invade in the orange space. (In the appendix we develop a simple
%approximation to the unstable equilibrium frequency
%\eqref{bistab_homozy_approx} that is also shown in  Supplementary Figure
%\ref{Bistab_homozyg_cost_fig}).
This unstable equilibrium frequency is high across much of
	parameter space, likely too high to be reached by drift, 
	and therefore the more restrictive invasion criteria rather than the fixation criteria, will  likely determine the fate of self promoting drivers. 
}

@@

Although the allelic identity of sperm could plausibly influence the outcome of female meiosis, 
	limited gene expression in sperm \citep[e.g.][]{Joseph2004}
	suggest a model where sperm influence female meiosis via expression of the fertilizing male's
	diploid genotype (perhaps due to proteins and RNAs packaged into the sperm), rather than sperm haplotype.
%GRAHAM I HAVE REWRITTEN THIS SECTION TO REFLECT MATHS AND MY PARAMETERIZATION. WE CAN WORK ON SYNCHING OUR NOTATION
%This adds an additional parameter to the system -- the dominance of the drive enhancer in heterozygous males, 
%	when this dominance is high, a narrow sliver of parameter space sustaining a polymorphism emerges (see Figure S\ref{Effect_of_dominance})
 %presumably because positive frequency dependence diminishes with the dominance of the male effect driver.
This male-genotype dependent model requires one additional parameter, as we exchange $d$ in the sperm dependent case for $d_{het}$ and $d_{hom}$ which describe the transmission of the drive allele in a heterozygous female mating with males heterozygous and homozygous for the self-promoting drive allele, respectively.  
When $d_{het}$ is large relative to $d_{hom}$, a narrow sliver of parameter space sustains a polymorphism (see Figure S\ref{Effect_of_dominance}), presumably because positive frequency dependence diminishes with the dominance of the male effect driver. 
As an analytical example, consider the case of male-genotype dependent deleterious recessive female driver 
	(assuming, as above, $w_{BB}=1-s$, and $w_{AA}=w_{AB}=1$). 
Here, a rare driver invades when $s$ is less than $(d_{het}-1/2)/d_{het}$, and fixes when $s$ is less than $d_{hom} - 1/2$. 
Intermediate values of $s$, which represent a small proportion of parameter space can maintain a polymorphism.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%
% TWO LOCUS MODEL [modify (general) / facilitate (helps ) / prevent (stops)] 
%%%%%%%%%%%%%%%%%%%%%%%%%
\yb{********* DO WE NEED TO DO THE MALE GENOTYPE THING HERE TOO? *********}

@@

\paragraph{Dominance} we also explored a model where
the genotype of the male who produced the fertilizing sperm affects female drive,
instead of just the allelic type of the sperm.
 We assume a heterozygous female will transmit the $B$ allele 
  with probabilities  $\frac{1}{2}$,  $\frac{1}{2} + h(d-\frac{1}{2}) $ or $d$, 
 if she mated with a n $AA$, $AB$, or $BB$ male,  respectively.
 Above, $d > 1/2$ and $h$ ($0 \geq h \geq 1$) acts to specify the dominance of the $B$
%allele's effect on female meiotic drive through the males genotype.
%YB TRY_1 We assume that in a mating between a heterozygous female and an $AA$, $AB$, or $BB$ male, the female will transmit the $B$ allele 
% with probabilities  $\frac{1}{2}$,  $\frac{1}{2} + h(d-\frac{1}{2}) $, and $d$, respectively.
  %Here $d > 1/2$ and $h$ ($0 \geq h \geq 1$) acts to specify the dominance of the $B$
%allele's effect on female meiotic drive through the males genotype.
% GC ORIGINAL  We assume that if a
%heterozygous egg ($AB$) is fertilized by a sperm with the following
%male ($\Mars$) genotypes the $B$ allele will be transmitted through
%female meiosis with the following probabilities \\
%\begin{tabular}{cccc}
%$\Mars$ geno. & AA & AB  & BB\\
%P(B). & $\frac{1}{2}$ &  $\frac{1}{2} + h(d-\frac{1}{2}) $ & $d$ 
%\end{tabular}\\
%where $d > 1/2$ and $h$ ($0 \geq h \geq 1$) acts to specify the dominance of the $B$
%allele's effect on female meiotic drive through the males genotype. 
We set up a set of recursions using a recessive sex independent cost to
the driver, similar to those in eqn.
\ref{eqn.A1}-\ref{eqn.A1}, but now featuring these male genotype
transmission probabilites. In Supplementary Figure \ref{Effect_of_dominance} we show the invasion
criteria for both the A and B alleles when they are rare. 
We find a narrow parameter range where B 
can invade but not fix.
This corresponds to B being less than fully recessive (i.e. $h>0$) and to a modest cost in homozygotes. 
This combination of parameters reflects the fact that when the allele is rare, 
it can increase in frequency due to fertilization of heterozygous eggs
by sperm from heterozygote males, where the sperm do not themselves
carry the allele. But as the B allele becomes common it will frequently be
present in the homozygous state in males, and so will be unable to
avoid the production of unfit homozygote zygotes leaving it trapped in
the population. \gc{Perhaps we should show the equil. freq. in the
  fig, as I'm guessing it might be rare}

@@

\begin{figure}
\includegraphics[width = 0.8 \textwidth]{../Scripts/effect_of_dominance_on_invasion_space_four_graph.eps} 
\caption{Invasion analysis of an allele whose effect
 on female meoisis is mediated by the genotype of the fertilizing
 male.  As above wee assume a heterozygous female will transmit the $B$ allele 
  with probabilities  $\frac{1}{2}$,  $\frac{1}{2} + h(d-\frac{1}{2}) $ or $d$, 
 if she mated with a n $AA$, $AB$, or $BB$ male,  respectively.  The allele suffers a recessive fitness cost $s$.  
 The four panels correspond to different dominance relationships.
In the parameter space below the invasion (solid) line the self promotor
 driver can invade. In the parameter space below the fixation (long
 dash) line the self promotor can fix in the population. In the last
 two panels the invasion line is above the fixation line and so the
 allele can be maintained as a polymorphism in that thin slice of
 parameter space between the two lines. These results were 
  obtained using the recursion. In the final panel we show the
  fixation line (small dashes) as predicted by our HWE approximation.
}  \label{Effect_of_dominance}
\end{figure}

>>>>

%I SUBSTANTIALLY REWROTE THIS
The positive frequency dependence of a self-promoting driver can
induce bistability in its spread.
%OLD GRAHAM WORDS
 %, i.e. there is a critical frequency
%above which it can spread quickly to fixation and below which it moves towards loss (an
%unstable equilibrium frequency). In 
%the appendix, we develop a simple
%approximation to this bistability. 
When $s$ is too large to allow a rare driver to increase in frequency, but small enough to allow a common driver to fix (the inequality in EQUATION XXX),
there is  an unstable equilibrium frequency (approximated in EQUATION XXX)  
	above which it can spread quickly to fixation and below which it moves towards loss 
	(see Figure \ref{Bistab_homozyg_cost_fig} for exact results).	
%OLD GRAHAM WORDS
%This bistability occurs in the region of parameter space in Figure
%\ref{Invasion_space}. In Supplementary Figure
%\ref{Bistab_homozyg_cost_fig} we further explore this bistability and
%show that the unstable equilibrium frequency is high across much of
%this parameter space, and so in practice the self promoting allele
%only invade in the orange space. (In the appendix we develop a simple
%approximation to the unstable equilibrium frequency
%\eqref{bistab_homozy_approx} that is also shown in  Supplementary Figure
%\ref{Bistab_homozyg_cost_fig}).
This unstable equilibrium frequency is high across much of
	parameter space, likely too high to be reached by drift, 
	and therefore the more restrictive invasion criteria rather than the fixation criteria, will  likely determine the fate of self promoting drivers.

@@

Although the allelic identity of sperm could plausibly influence the outcome of female meiosis, 
	limited gene expression in sperm \citep[e.g.][]{Joseph2004}
	suggest a model where sperm influence female meiosis via expression of the fertilizing male's
	diploid genotype (perhaps due to proteins and RNAs packaged into the sperm), rather than sperm haplotype.
%GRAHAM I HAVE REWRITTEN THIS SECTION TO REFLECT MATHS AND MY PARAMETERIZATION. WE CAN WORK ON SYNCHING OUR NOTATION
%This adds an additional parameter to the system -- the dominance of the drive enhancer in heterozygous males, 
%	when this dominance is high, a narrow sliver of parameter space sustaining a polymorphism emerges (see Figure S\ref{Effect_of_dominance})
 %presumably because positive frequency dependence diminishes with the dominance of the male effect driver.
This male-genotype dependent model requires one additional parameter, as we exchange $d$ in the sperm dependent case for $d_{het}$ and $d_{hom}$ which describe the transmission of the drive allele in a heterozygous female mating with males heterozygous and homozygous for the self-promoting drive allele, respectively.  
However, when
the distorting effect of genotype displays strong dominance in its
effect on female meiosis ($d_{het}$ is close to $d_{hom}$), 
there is a a narrow sliver of parameter space sustains a polymorphism
when the cost of the drive is recessive
(see Figure S\ref{Effect_of_dominance}). 
Presumably this zone of polymorphism comes about because positive frequency dependence diminishes with the dominance of the male effect driver. 
As an analytical example, consider the case of male-genotype dependent deleterious recessive female driver 
	(assuming, as above, $w_{BB}=1-s$, and $w_{AA}=w_{AB}=1$). 
Here, a rare driver invades when $s$ is less than $(d_{het}-1/2)/d_{het}$, and fixes when $s$ is less than $d_{hom} - 1/2$.
Intermediate values of $s$, which represent a small proportion of parameter space can maintain a polymorphism. 
However, it does not seem particularly realistic to think that the
effect of the drive allele would be dominant in its action through the
male genotype, while the cost would be recessive. Thus we regard this
case as interesting but not very general.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%
% TWO LOCUS MODEL [modify (general) / facilitate (helps ) / prevent (stops)] 
%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\paragraph{Dominance} we also explored a model where
the genotype of the male who produced the fertilizing sperm affects female drive,
instead of just the allelic type of the sperm.
 We assume a heterozygous female will transmit the $B$ allele 
  with probabilities  $\frac{1}{2}$,  $\frac{1}{2} + h(d-\frac{1}{2}) $ or $d$, 
 if she mated with a n $AA$, $AB$, or $BB$ male,  respectively.
 Above, $d > 1/2$ and $h$ ($0 \geq h \geq 1$) acts to specify the dominance of the $B$
%allele's effect on female meiotic drive through the males genotype.
%YB TRY_1 We assume that in a mating between a heterozygous female and an $AA$, $AB$, or $BB$ male, the female will transmit the $B$ allele 
% with probabilities  $\frac{1}{2}$,  $\frac{1}{2} + h(d-\frac{1}{2}) $, and $d$, respectively.
  %Here $d > 1/2$ and $h$ ($0 \geq h \geq 1$) acts to specify the dominance of the $B$
%allele's effect on female meiotic drive through the males genotype.
% GC ORIGINAL  We assume that if a
%heterozygous egg ($AB$) is fertilized by a sperm with the following
%male ($\Mars$) genotypes the $B$ allele will be transmitted through
%female meiosis with the following probabilities \\
%\begin{tabular}{cccc}
%$\Mars$ geno. & AA & AB  & BB\\
%P(B). & $\frac{1}{2}$ &  $\frac{1}{2} + h(d-\frac{1}{2}) $ & $d$ 
%\end{tabular}\\
%where $d > 1/2$ and $h$ ($0 \geq h \geq 1$) acts to specify the dominance of the $B$
%allele's effect on female meiotic drive through the males genotype. 
We set up a set of recursions using a recessive sex independent cost to
the driver, similar to those in eqn.
\ref{eqn.A1}-\ref{eqn.A1}, but now featuring these male genotype
transmission probabilites. In Supplementary Figure \ref{Effect_of_dominance} we show the invasion
criteria for both the A and B alleles when they are rare. 
We find a narrow parameter range where B 
can invade but not fix.
This corresponds to B being less than fully recessive (i.e. $h>0$) and to a modest cost in homozygotes. 
This combination of parameters reflects the fact that when the allele is rare, 
it can increase in frequency due to fertilization of heterozygous eggs
by sperm from heterozygote males, where the sperm do not themselves
carry the allele. But as the B allele becomes common it will frequently be
present in the homozygous state in males, and so will be unable to
avoid the production of unfit homozygote zygotes leaving it trapped in
the population.

@@

\begin{figure}
\includegraphics[width = 0.8 \textwidth]{../Scripts/effect_of_dominance_on_invasion_space_four_graph.eps} 
\caption{Invasion analysis of an allele whose effect
 on female meoisis is mediated by the genotype of the fertilizing
 male.  As above wee assume a heterozygous female will transmit the $B$ allele 
  with probabilities  $\frac{1}{2}$,  $\frac{1}{2} + h(d-\frac{1}{2}) $ or $d$, 
 if she mated with a n $AA$, $AB$, or $BB$ male,  respectively.  The allele suffers a recessive fitness cost $s$.  
 The four panels correspond to different dominance relationships.
In the parameter space below the invasion (solid) line the self promotor
 driver can invade. In the parameter space below the fixation (long
 dash) line the self promotor can fix in the population. In the last
 two panels the invasion line is above the fixation line and so the
 allele can be maintained as a polymorphism in that thin slice of
 parameter space between the two lines. These results were 
  obtained using the recursion. In the final panel we show the
  fixation line (small dashes) as predicted by our HWE approximation
  ($(d_{het}-1/2)/d_{het}$), see the `dominance' section of the
  Appendix for more details.
}  \label{Effect_of_dominance}
\end{figure}
